### 2017-02-05 doc/manuscript.Rmd
Edits as per Tony's suggestions from doc # Create branch

(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

@@

Hypertriglyceridemia is an extensively studied and well known factor involved in
the dysregulation of metabolic function and subsequent negative health impacts.
It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
cardiovascular disease [@DAgostino2004a], and can contribute to the pathogenesis
of diabetes. Measuring triacylglcerides (TAG) within the blood is commonly done
during routine clinical visits as part of health assessment. However, clinically
measured TAG is limited as it represents all the types of fatty acids within the
TAG fraction as a single measure. Given the increasing appreciation for the role
of the specific fatty acid composition on health outcomes, for instance with
differences in associations of fatty acids within the PL lipid fraction
[@Forouhi2014a; @Ma2015a] including an analysis from the PROMISE cohort
[@Johnston2016a], there are relatively few studies that have explored this area
of research in the TAG fraction.

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system. Greater resistance to insulin in both the liver
and muscle may result in greater production of TAG and secretion of lipoproteins
that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
metabolic dysfunction and lipotoxicity in various tissues, such as the 
beta-cells, and thus continuing the cycle. Given the complexity and temporal
nature of the relationship, long term studies with multiple data collections are
paramount to better understanding the underlying biology and subsequent risk.

While several prospective studies have identified a harmful association of 
hypertriglyceridemia on greater incidence for type 2 diabetes 
[@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
studies that have examined the relationship between TAG and it's composition on
the pathophysiological factors underlying diabetes, particularly in regard to
beta-cell function. Our objective was to examine the longitudinal role of the
specific composition of the serum TAG fraction on OGTT-derived measures of
insulin sensitivity and beta-cell function in an Canadian population at risk for
diabetes.

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
collected at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples taken during fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Questionnaires administered at each examination determined sociodemographics.
A version of the Modifiable Activity Questionnaire (MAQ) [@Kriska1990]
determined estimated physical activity. The MAQ collects information on leisure
and occupational activity, including intensity, frequency, and duration, over
the past year. Each reported activity from the MAQ was weighted by its
metabolic intensity allowing for the estimation of MET-hours per week.

@@

The primary outcome variables for this analysis were HOMA2-%S, ISI, IGI/IR, and
ISSI-2, which were log-transformed for the statistical modeling. The primary
predictor variables for this analysis were 22 individual TAGFA as mole percent
(mol%) of the total fraction and as concentration (nmol/mL). Clinically-measured
TAG is also included as a primary predictor to serve as a comparison against the
TAGFA composition. A limitation with the clinically-measured TAG is that it was
only quantified on a portion of the participants at the 6-year visit (n = `r n_tag`).
Pearson correlation coefficients were computed to assess the relationships of
individual TAGFA with other continuous variables. Correlations were also
computed for TAGFA against each other, followed by using hierarchical clustering
analysis to identify clusters within the correlation matrix of the TAGFA
composition.

@@

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
environment (all TAGFA), partial least squares regression (PLS) was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function. No prevalent diabetes cases were included in the PLS
analysis. For more detailed explanation of PLS, please see the Supplemental
Methods.

@@

`r cite_t('basic')` shows some basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)`. Given the eligibility criteria, the majority of
participants (`r fhd$npct`) had a family history of diabetes.
Between the baseline visit and the 6-year visit in this sample, insulin
sensitivity and beta-cell function measures had a significant median decline
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$dm` and `r dys$pre` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively.

@@

Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
for the adjusted GEE model in `r cite_f('gee_adj')`. 
The majority of associations with beta-cell function measures were attenuated
after full model adjustment, while nearly all associations with insulin
sensitivity remained significant. 
<!--{{ confirm }} Subsequent analysis revealed
that the attenuation with beta-cell function was due primarily to adjustment for
{{ waist? ALT? }}.-->

As a concentration, nearly all TAGFA had a strong negative association on
HOMA2-%S and ISI (estimates ranging from 
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase in 
20:2n-6.

@@

Confirming the findings from the GEE models, PLS analysis on the beta-cell
function measures revealed that there was a low correlation (though significant
at p`r cor_pls_igi$p`) between the predicted values and the actual values (r=
`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability. While 
the clustering of TAGFA from the PLS results of the beta-cell function measures 
were similar to the insulin sensitivity measures, only the insulin sensitivity 
measures are presented given the higher predictive ability.

The PLS results for the insulin sensitivity (`r cite_f('pls')`) revealed a clustering 
of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
on both insulin sensitivity measures. No other TAGFA loaded strongly. The PLS
model for insulin sensitivity had good predictive ability, determined by
correlating the predicted outcome values with the actual values (HOMA2-%S: `r
cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

We found that in an at-risk for diabetes Canadian population who were mainly female 
and of European ancestry, the specific TAGFA composition was strongly associated 
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TAGFA were not
associated with beta-cell function, two fatty acids, 16:0 and 18:1n-7, were
associated negatively and positively, respectively, with both measures of 
beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
clustered together, and that also strongly predicted lower insulin sensitivity. 
These four fatty acids are also involved in the DNL of refined and simple 
carbohydrates. Our results, which are in agreement with the current literature,
suggest that higher activity of DNL (potentially through higher intakes of
simple carbohydrates) may increase the risk for diabetes, likely through
worsening insulin sensitivity.

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? appropriate?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
study [@Rhee2011a], which found that TAGFA composition with a lower carbon chain
and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
the study was the simpler statistical techniques used to analyze the data, 
which are not able to adequately use the longitudinal and high-dimensional 
dataset. {{ Their article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
report a positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings on DNL by showing that TAGFA with 14 to 16 carbons all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A short, small feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet,
particularly with simple carbohydrates, may lead to greater DNL in an attempt to
control blood glucose, increasing hepatic fat stores, and increasing the amount
of TAGFA in circulation. The higher concentration of circulating 14 and 16
carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
from palmitic acid (16:0), which is well-known to have harmful effects on
tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
null results on this question may tentatively point to an equally influential
feedback mechanism between TAGFA and insulin sensitivity.

>>>>

(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
(3) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

@@

Hypertriglyceridemia is an extensively studied and well known factor involved in
the dysregulation of metabolic function and subsequent negative health outcomes {{cite}}.
It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
cardiovascular disease [@DAgostino2004a], and has been shown to contribute to the pathogenesis
of diabetes {{cite}}. The measurement of circulating triacylglcerides (TAG) is commonly done
during routine clinical visits as part of health assessment. However, clinically
measured TAG is limited as it represents all the types of fatty acids within the
TAG fraction as a single measure. Given the increasing appreciation for the role
of the specific fatty acid composition on health outcomes, for instance with
differences in associations of fatty acids within the PL lipid fraction
[@Forouhi2014a; @Ma2015a] including an analysis from the PROMISE cohort
[@Johnston2016a], there are relatively few studies that have explored this area
of research in the TAG fraction.

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system {{cite}}. Greater resistance to insulin in both the liver
and muscle may result in greater production of TAG and secretion of lipoproteins
that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
metabolic dysfunction and lipotoxicity in various tissues, such as the 
beta-cells, and thus continuing the cycle {{cite}}. Given the complexity and temporal
nature of the relationship, long term studies with multiple data collections are
paramount to better understanding the underlying biology and subsequent risk.

While several prospective studies have documented a prospective association of 
hypertriglyceridemia with incident type 2 diabetes 
[@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
studies that have examined the relationship between TAG and it's composition on
the pathophysiological factors underlying diabetes, particularly in regard to
beta-cell function. Our objective was to examine the longitudinal role of the
specific composition of the serum TAG fraction on OGTT-derived measures of
insulin sensitivity and beta-cell function compared to clinically measured TAG
in an Canadian population at risk for diabetes.

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 in London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
administered at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

After 8-12 hours of fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples taken at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured TAG 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON). At the 6-year visit only, clinically-measured TAG was only quantified on a
subset of the participants (n=`r n_tag`). Both insulin and glucose were
measured from OGTT blood samples at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Questionnaires administered at each examination determined sociodemographics.
A version of the Modifiable Activity Questionnaire (MAQ) [@Kriska1990]
determined estimated physical activity. The MAQ collects information on leisure
and occupational activity, including intensity, frequency, and duration, over
the past year. Each reported activity from the MAQ was weighted by its
metabolic intensity allowing for the estimation of MET-hours per week [@Kriska1990].

@@

The primary outcome variables for this analysis were HOMA2-%S, ISI, IGI/IR, and
ISSI-2, which were log-transformed for the statistical modeling. The primary
predictor variables for this analysis were 22 individual TAGFA included as either mole percent
(mol%) of the total fraction or as a concentration (nmol/mL). Clinically-measured
TAG was also included as a primary predictor to allow us to test the hypothesis
that specific TAGFA better predicted outcomes compared to clinical TAG. 
Pearson correlation coefficients were computed to assess the relationships of
individual TAGFA with other continuous variables. Correlations were also
computed for TAGFA against each other, followed by using hierarchical clustering
analysis to identify clusters within the correlation matrix of the TAGFA
composition.

@@

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function. For a detailed explanation of PLS see the Supplemental
Methods, but briefly, PLS is a technique that extracts latent structures
(clusters) underlying a set of predictor variables conditional on a response
variable(s) (i.e. the outcome variables). These extracted latent structures can
be used to calculated the predicted value of the outcome values, determining how
well TAGFA predict metabolic function by comparing to the measured metabolic function.

@@

`r cite_t('basic')` shows basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()` years, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility criteria, the majority of
participants (`r fhd$npct`) had a family history of diabetes.
Between the baseline visit and the 6-year visit in this sample, insulin
sensitivity and beta-cell function measures had a significant median decline
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$dm` and `r dys$pre` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years, of whom were excluded from later analyses.

@@

Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
for the adjusted GEE model in `r cite_f('gee_adj')`. The majority of
associations with beta-cell function measures were attenuated after full model
adjustment, while nearly all associations with insulin sensitivity remained
significant for both mol% and nmol/mL results. Subsequent analysis revealed that
the attenuation with beta-cell function was due primarily to adjustment for
waist.

In analyses using TAGFA concentration values, nearly all TAGFA had a strong
negative association on HOMA2-%S and ISI (estimates ranging from
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase in 
20:2n-6.

@@

The PLS analysis corroborated the findings from the GEE models. 
The insulin sensitivity (`r cite_f('pls')`) PLS results showed a clustering 
of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly
and negatively on HOMA2-%S and ISI, suggesting this cluster of TAGFA tracks
together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
TAGFA loaded strongly. The PLS model for insulin sensitivity had good predictive
ability, with a high correlation between the predicted outcome values against
the actual values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

The beta-cell function PLS results showed clustering of similar fatty acids,
however there there was a low correlation (though significant at 
p`r cor_pls_igi$p`) between the predicted values and the actual values 
(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability.
Given the low predictability, only the insulin sensitivity measures are
presented.

@@

We found that in an at-risk for diabetes Canadian population who were mainly female 
and of European ancestry, the specific TAGFA composition was strongly associated 
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TAGFA were not
associated with beta-cell function, two fatty acids, palmitic acid (16:0) and *cis*-vaccenic acid (18:1n-7), were
associated negatively and positively, respectively, with both measures of 
beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
clustered together, and this cluster also strongly predicted lower insulin sensitivity. 
These four fatty acids are also involved in the *de novo* lipogenesis (DNL) of refined and simple 
carbohydrates [@Hodson2008a; @Kawano2013a]. Our results, which are in agreement with the current literature,
suggest that higher activity of DNL (potentially through higher intakes of
simple carbohydrates) may increase the risk for diabetes, likely through
worsening insulin sensitivity.

@@

To our knowledge, no longitudinal study to date has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? maybe 'appropriate'?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control analysis (n=189 cases and n=189 controls) within the Framingham offspring
cohort [@Rhee2011a], which found that TAGFA composition characterized by a lower carbon chain
and fewer double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. As with our results, they found a larger effect size for
insulin sensitivity compared to the beta-cell function. There were some
similarities in results between the present analysis and the Finnish study,
specifically for the saturated fatty acids, docosapentaenoic acid (22:5n-3),
eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
acid (16:1n-7). Our study extends these findings by using multiple measurements
of metabolic function and using statistical techniques that adequately make use
of the longitudinal and high-dimensional dataset. In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TAGFA in circulation {{cite}}. The higher
concentration of circulating 14 and 16 carbon fatty acids may then expose
tissues to greater lipotoxicity, for instance from palmitic acid (16:0), which
is well-known to have harmful effects on tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
null results on this question may tentatively point to an equally influential
feedback mechanism between TAGFA and insulin sensitivity {{rephrase after exploring time interaction}}.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(under review), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
composition such as dietary intake {{cite?}}.

<!--
{{ Not sure about this... Include this? }}
As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
that cis-vaccenic acid strongly predicted higher insulin sensitivity and
beta-cell function. Likewise, in the TAGFA fraction we found a similar
association with cis-vaccenic acid. {{we can discuss to include more}}

{{ comment on comparing models with and without NEFA }} While
NEFA may be a mediating factor, we found that adjustment for it did not
{{confirm}} attenuate this association since TAGFA composition is also dictated
by other sources for DNL (e.g. recent meal, DNL from glucose stores, etc.).
